Subsidiary Recursion in Coq

### 2022-01-10 paper/paper.tex
further text

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
  This paper describes a functor-generic derivation of a recursion
  scheme supporting the initiation of subsidiary recursions, in Coq.
  The approach supports making outer recursive calls on the results of
  subsidiary recursions.  The derivation utilizes a novel (necessarily
  weakened) form of positive-recursive types in Coq.  Subsidiary
  induction is also supported.  Using this derivation, several
  examples are demonstrated.
\end{abstract}

@@

Central to interactive theorem provers like Coq, Agda, Isabelle/HOL,
Lean and others are terminating recursive functions over user-declared
inductive datatypes~\cite{agda,coq,isabelle-hol,lean}.  Termination is
usually enforced by a syntactic check for structural decrease.  This
structural termination is sufficient for many basic functions.  For
example, the well-known \texttt{span} function from Haskell's standard
library (\verb|Data.List|) takes a list and returns a pair of the
maximal prefix satisfying a given predicate \verb|p|, and the
remaining suffix:
\begin{verbatim}
span :: (a -> Bool) -> [a] -> ([a],[a])
span _ []    =  ([], [])
span p (x:xs) = if p x
                then let (ys,zs) = span p xs in (x:ys,zs)
                else ([],x:xs)
\end{verbatim}
\noindent The sole recursive call is \verb|span p xs|, and it occurs
in a clause where the input list is of the form \verb|x:xs|.  The
input to the recursive call is a subdatum of the input for that
clause, and hence this definition is structurally decreasing.  In the
appropriate syntax, it can be accepted without additional effort by
all the mentioned provers.

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further recursion.  Let us see a simple example.  The
function \verb|wordsBy| (\verb|Data.List.Extra|) breaks a list into
the list of its maximal sublists whose elements do not satisfy a
predicate \verb|p|.  For example, \verb|wordsBy isSpace " good day "|
returns \verb|["good","day"]|; so \verb|wordsBy isSpace| has the same
behavior as \verb|words| (\verb|Data.List|).  Code is in
Figure~\ref{fig:wordsBy}. The first recursive call,
\verb|wordsBy p tl|, is structural.  But in the second, we invoke
\verb|wordsBy p| on a value obtained from another recursion, namely
\verb|span|.  This is not allowed under structural termination.

@@

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We should emphasize that
this is a derivation of this recursion scheme within the type theory
of Coq.  (So no axioms or other modifications to Coq of any kind are
required.) Based on this derivation, we present several example
functions like \verb|wordsBy|, and prove theorems about them.  For
example, we prove the expected property that the sublists returned by
\verb|wordsBy| contain no elements satisfying the predicate \verb|p|.
For another, we give a definition of run-length encoding as a
subsidiary recursion using \verb|span|, and prove that encoding and
then decoding returns the original list.  The approach applies to the
standard datatypes in the Coq library, and does not require switching
libraries or datatype definitions.

An important technical novelty of our approach is a derivation of a
weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
restrict datatypes $D$ to be strictly positive: in the type for any
constructor of $D$, $D$ cannot occur to the left of any arrows.  Our
derivation needs to use positive-recursive types, where $D$ may occur
to the left only of an even number of arrows.  Full positive-recursive
types are incompatible with Coq's type theory.  But we present a way
to derive a weakened form that is sufficient for our examples
(Section~\ref{sec:mu}).  The weakening is to require just a retraction
between recursive type $\mu$ and its one-step unfolding $F\ \mu$,
instead of an isomorphism.  Hence, we dub these
\textbf{retractive-positive} recursive types.

We begin by summarizing the interface our derivation provides for
subsidiary recursion (Section~\ref{sec:interface}), and then see
examples (Section~\ref{sec:examples}).  The interface for subsidiary
induction is covered next (Section~\ref{sec:interfacei}), and example
proofs using it (Section~\ref{sec:examplesi}).  We conclude with a
walk-through of the derivation of that first interface (for subsidiary
recursion), in Section~\ref{sec:deriv}.  Related work is discussed in
Section~\ref{sec:related}.

All presented derivations have been checked with Coq version 8.13.2.
The code may be found as release \verb|itp-2022| (dated prior to the
ITP 2022 deadline) at
\url{https://github.com/astump/coq-subsidiary}.

\section{Interface}
\label{sec:interface}

@@

Definition inList : ListF List -> List := inn ListF.
Definition outList : List -> ListF List := out ListF (fold ListF).
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{verbatim}
  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists}
  \label{fig:listf}
\end{figure}

For purposes of presenting the interface for subsidiary recursion, let us consider three files in our codebase:

\begin{itemize}
\item \texttt{Subrec.v}: parametrized by a signature functor \verb|F| of type \verb|Set -> Set|,
  this provides (among much else, to be discussed below) \verb|Subrec : Set|, as the type to use
  for subsidiary recursion; and also a constructor \verb|inn : F Subrec -> Subrec| for that type.

\item \texttt{List.v}: parametrized by the type \verb|A : Set| of list
  elements, this file specializes the development in \verb|Subrec.v|
  to the case of lists.  This is done using defining the signature
  functor \verb|ListF|, shown in Figure~\ref{fig:listf}, giving us the
  type \verb|List|.  This is not to be confused with the type
  \verb|list| of lists in Coq's standard library.  The figure also
  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
  and types for conversion functions between \verb|List| and
  \verb|list| (code omitted).

\item \texttt{Span.v}: this implements the \verb|span| function
  mentioned above, as an \emph{algebra}.

\end{itemize}

@@

Our approach is within a long line of work using ideas from
universal algebra and category theory to implement inductive datatypes
and recursion principles, in type theory.  On this approach, one describes
transformations to be performed on data as \emph{algebras}, which can then
be \textit{folded} over data.

@@

Definition Alg : KAlg := MuAlg AlgF.
\end{verbatim}
\caption{The interface for algebras}
\label{fig:algf}
\end{figure}

The interface for algebras is presented in Figure~\ref{fig:algf}.
Let us consider the definitions.  \verb|KAlg| is the kind for the type-constructor for algebras,
as we see in the very last definition of the figure, for \verb|Alg|.  This type-constructor \verb|Alg|
is a fixed-point of the type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|,
to be discussed in Section~\ref{sec:mu} below.  Doing so requires that \verb|AlgF| only use its parameter
\verb|Alg| positively.  We will confirm this shortly.

The type \verb|FoldT Alg C| is the type for fold functions which apply
algebras of type \verb|Alg| to data of type \verb|C|.  At the top
level of code, \verb|C| would just be \verb|List| (for example).  When
one initiates a subsidiary recursion, though, this type \verb|C| will
instead by what we like to call the \emph{recursion universe} of that
recursion.  Each recursion is based on an abstract type \verb|R|,
representing the data upon which we will recurse.  Indeed, \verb|AlgF|
says that algebras take in such an \verb|R| as their first argument.
This means that an algebra can assume nothing about \verb|R| except
that it has the operations given next in the listing of \verb|AlgF|.
Thus, \verb|R| functions like a universe for recursion: one may
perform the various given operations, without leaving the type
\verb|R| (which is good, because we may recurse on elements of type \verb|R|).

And what are the operations on \verb|R|?  First there is \verb|reveal|, which
turns an \verb|R| into a \verb|C|.  This reveals that the data of type
\verb|R| are really lists (if this is a top-level recursion) or else
really belong to some outer recursion universe (if this is an inner
recursion).  Next we have \texttt{fold}, which will allow us to fold
another algebra over data of type \verb|R|.  We will use \verb|fold|
to initiate subsidiary recursions.  Then there is \verb|eval|, which
is used to make recursive calls on data of type \verb|R|.

This is a good place to highlight that for subsidiary recursion, it is
crucial that algebras have a carrier \verb|X| which depends
(functorially) on a type.  This is so that (i) inside an inner
recursion we may compute a result of some type that may mention
\verb|R|, but (ii) outside that recursion, the result will mention the
type \verb|C|.  The \verb|eval| function returns something of type
\verb|X R|, demonstrating (i).  For (ii): if we look at the definition
of \verb|FoldT| in the figure, we see that folding an algebra of type
\verb|alg C X| over a value of type \verb|C| produces a result of type
\verb|X C|.

@@

\section{Examples of subsidiary induction}
\label{sec:interfacei}

@@

@@

In some tools, like Coq, Agda, and Lean, termination is checked
statically, based on structural decrease at recursive calls.  Others,
like Isabelle/HOL, allow one to write recursions first, and prove
(possibly with automated help) their termination
afterwards~\cite{krauss}.

@@

Subsidiary recursion can be seen as a generalization
of nested recursion, which is a form of recursion allowing recursive
calls of the form \verb|f\ (f\ x)|.  In subsidiary recursion, these
are generalized to the form \verb|f\ (g\ x)|, where \verb|g| is
another recursively defined function.

>>>>

\newcommand{\all}[2]{\forall\, #1.\, #2}

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
  This paper describes a functor-generic derivation in Coq of
  subsidiary recursion.  On this recursion scheme, inner recursions
  may be initiated within outer ones, in such a way that outer
  recursive calls may be made on results from inner ones.  The
  derivation utilizes a novel (necessarily weakened) form of
  positive-recursive types in Coq, dubbed retractive-positive
  recursive types.  A corresponding form of induction is also
  supported.  The method is demonstrated through several examples.
\end{abstract}

@@

Central to interactive theorem provers like Coq, Agda, Isabelle/HOL,
Lean and others are terminating recursive functions over user-declared
inductive datatypes~\cite{agda,coq,isabelle-hol,lean}.  Termination is
usually enforced by a syntactic check for structural decrease.  This
structural termination is sufficient for many basic functions.  For
example, the well-known \texttt{span} function from Haskell's standard
library (\verb|Data.List|) takes a list and returns a pair of the
maximal prefix satisfying a given predicate \verb|p|, and the
remaining suffix:
\begin{verbatim}
span :: (a -> Bool) -> [a] -> ([a],[a])
span _ []    =  ([], [])
span p (x:xs) = if p x
                then let (ys,zs) = span p xs in (x:ys,zs)
                else ([],x:xs)
\end{verbatim}
\noindent The sole recursive call is \verb|span p xs|, and it occurs
in a clause where the input list is of the form \verb|x:xs|.  So the
input to the recursive call is a subdatum of the input, and hence this
definition is structurally decreasing.  In the appropriate syntax, it
can be accepted without additional effort by all the mentioned
provers.

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further invocations of the outer recursive function.
Let us see a simple example.  The function \verb|wordsBy|
(from \verb|Data.List.Extra|) breaks a list into its maximal
sublists whose elements do not satisfy a predicate \verb|p|.  For
example, \verb|wordsBy isSpace " good day "| returns
\verb|["good","day"]|; so \verb|wordsBy isSpace| has the same behavior
as \verb|words| (from \verb|Data.List|).  Code is in
Figure~\ref{fig:wordsBy}. The first recursive call,
\verb|wordsBy p tl|, is structural.  But in the second, we invoke
\verb|wordsBy p| on a value obtained from another recursion, namely
\verb|span|.  This is not allowed under structural termination,
but will be permitted by subsidiary recursion as derived below.

@@

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We should emphasize that
this is a derivation of this recursion scheme within the type theory
of Coq.  No axioms or other modifications to Coq of any kind are
required. Based on this derivation, we present several example
functions like \verb|wordsBy|, and prove theorems about them.  For
example, we prove the expected property that the sublists returned by
\verb|wordsBy| consist of elements satisfying \verb|not . p|.
For another, we give a definition of run-length encoding as a
subsidiary recursion using \verb|span|, and prove that encoding and
then decoding returns the original list.  Our approach applies to the
standard datatypes in the Coq library, and does not require switching
libraries or datatype definitions.

An important technical novelty of our approach is a derivation of a
weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
restrict datatypes $D$ to be strictly positive: in the type for any
constructor of $D$, $D$ cannot occur to the left of any arrows.  Our
derivation needs to use positive-recursive types, where $D$ may occur
to the left only of an even number of arrows.  The restriction to
strict positivity is enforced because full positive-recursive types
are incompatible with Coq's type theory~\cite{coquand88}.  But we
present a way to derive a weakened form that is sufficient for our
examples (Section~\ref{sec:mu}).  The weakening is to require only
that $F\ \mu$ is a retract of $\mu$, where $\mu$ is the recursive type
and $F\ \mu$ its one-step unfolding.  Usually these types are
isomorphic.  Hence, we dub these \textbf{retractive-positive}
recursive types.  This weakening does have some negative consequences,
but we will see how to handle them.  Our definition of
retractive-positive recursive types makes essential use of impredicate
quantification, and hence cannot be soundly recapitulated in a
predicative theory like Agda's.

We begin by summarizing the interface our derivation provides for
subsidiary recursion (Section~\ref{sec:interface}), and then see
examples (Section~\ref{sec:examples}).  We next explain how the
interface is actually implemented (Section~\ref{sec:deriv}), including
our retractive-positive recursive types (Section~\ref{sec:mu}).  The
interface for subsidiary induction is covered next
(Section~\ref{sec:interfacei}), and example proofs using it
(Section~\ref{sec:examplesi}).  Related work is discussed in
Section~\ref{sec:related}.

All presented derivations have been checked with Coq version 8.13.2,
using command-line option \verb|-impredicative-set|.  The code may be
found as release \verb|itp-2022| (dated prior to the ITP 2022
deadline) at \url{https://github.com/astump/coq-subsidiary}.

\section{Interface for subsidiary recursion}
\label{sec:interface}

This section presents the interface our Coq development provides
for subsidiary recursion.

\subsection{The recursion universe}

Our approach is within a long line of work using ideas from universal
algebra and category theory to describe inductive datatypes and their
recursion principles.  On this approach, one describes transformations
to be performed on data as \emph{algebras}, which can then be
\textit{folded} over data.  The simplest form of algebras, namely
$F$-algebras, are morphisms from $F\ A$ to $A$, for carrier object
$A$.  From a programming perspective, an $F$-algebra is presented
with $F\ A$, and must compute a value of type $A$.

Algebras for our subsidiary recursion are quite a bit more complex than
this.  First, for reasons we will explain further below, the carrier
of the algebra will be an \verb|X : Set -> Set|.  Second, instead of
being given just $F\ A$ as input, the algebra is presented with:

\begin{itemize}
\item a type \verb|R : Set|
\item a function \verb|reveal : R -> C|, which reveals values of type \verb|R| as really having \emph{anchor type} \verb|C|.  This type \verb|C| will either by the abstract type of some outer recursion, or else our development's version of the actual datatype one is recursing over.
\item a function \verb|fold : FoldT Alg R|, which allows one to initiate subsidiary recursions in which the anchor type is \verb|R|.  We will present its type \verb|FoldT Alg R| below.
\item a function \verb|eval : R -> X R|, to use for making recursive calls, on any value of type \verb|R|.
\item and an \emph{subdata structure} \verb|d : F R|, where \verb|F| is the signature functor for the datatype.
\end{itemize}

\noindent The algebra is then required to produce a value of type \verb|F R|.

We use Coq inductive types for the signature functors \verb|F| of
various datatypes, thus enabling recursions to use Coq's
pattern-matching on the subdata structure \verb|d|.  So the style of
coding against this interface retains a similar feel to structural
recursions.  Unlike with structural termination, though, the interface
here is type-based and hence compositional.  As we will see, it
supports nested and higher-order recursions.

As in previous work, we dub this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have an entity (here, \verb|R|) from which one cannot
escape by using the available operations (for other cases:
$\epsilon_0$ and $\omega^-$, the physical universe and traveling at
the speed of light).  Staying in the recursion universe is good,
because we may recurse (via \verb|eval|) on any value of type
\verb|R|.

Some points must still be explained, particularly why \verb|X| has type \verb|Set -> Set|,
and the definition of \verb|FoldT|.  Let us see these details next.

\subsection{The interface in more detail}

Let us consider two files from our development.

@@

Definition inList : ListF List -> List := inn ListF.
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{verbatim}
  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists (\texttt{List.v})}
  \label{fig:listf}
\end{figure}

\subsubsection{Subrec.v}

  This file is parametrized by a signature functor \verb|F| of type
  \verb|Set -> Set|.  It provides the implementation of subsidiary
  recursion.  Two crucial values are \verb|Subrec : Set|, which is the
  type to use for subsidiary recursion; and
  \verb|inn : F Subrec -> Subrec|, which is to be used as a
  constructor for that type.  An important point, however, is that
  \verb|Subrec.v| does not provide an induction principle based on
  \verb|inn|.  Induction is derived later
  (Section~\ref{sec:interfacei}). \verb|Subrec.v| makes critical use
  of retractive-positive recursive types, to take a fixed-point based
  on \verb|F|.  We will present the definition of those in
  Section~\ref{sec:mu}.

  \subsubsection{List.v}

  This file specializes the development in \verb|Subrec.v| to the case
  of lists (parametrized by the type \verb|A| of elements).  In
  general, to use our development to get subsidiary recursion over
  some datatype, one will have a similar ``shim'' file.  The file
  defines the signature functor \verb|ListF|, shown in
  Figure~\ref{fig:listf}.  Using \verb|Subrec|, we then get a type
  \verb|List|.  This is not to be confused with the type \verb|list|
  of lists in Coq's standard library.  As noted previously, our
  development is meant to be used in extension of existing inductive
  datatypes, not replacing them.  The figure also shows constructors
  \verb|mkNil| and \verb|mkCons| for \verb|List|, and types for
  conversion functions between \verb|List| and \verb|list| (see
  Section~\ref{sec:deriv} for code).

@@

\verb|Subrec.v| also defines the notion of algebra that is used for writing
recursions.  The central definitions are in Figure~\ref{fig:algf}.
\verb|KAlg| is the kind for the type-constructor for algebras, as we
see in the very last definition of the figure, for \verb|Alg|.  This
type-constructor \verb|Alg| is a fixed-point of the type \verb|AlgF|.
The fixed-point is taken using \verb|MuAlg| (Section~\ref{sec:mu}
below).  Doing so requires that \verb|AlgF| only use its parameter
\verb|Alg| positively.  We will confirm this shortly.

@@

Definition Alg : KAlg := MuAlg AlgF.
\end{verbatim}
\caption{The type for algebras (\texttt{Subrec.v})}
\label{fig:algf}
\end{figure}

The type \verb|FoldT Alg C| is the type for fold functions which apply
algebras of type \verb|Alg| to data of type \verb|C|, which we have
already dubbed the \emph{anchor type} of the recursion.  At the top
level of code, the anchor type would just be \verb|List| (for
example).  When one initiates a subsidiary recursion, though, the
anchor type will instead by the abstract type \verb|R| for the outer recursion.

The variable \verb|Alg| occurs only positively (but not strictly
positively) in \verb|AlgF|, because it occurs negatively in
\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg C X|.  So
we can indeed take a fixed-point of \verb|AlgF| to define the constant
\verb|Alg|.

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg C X|.  An algebra can assume nothing about \verb|R|
except that it supports the following operations.  First there is
\verb|reveal|, which turns an \verb|R| into a \verb|C|.  This reveals
that the data of type \verb|R| are really values of the datatype
(\verb|List|, for example), if this is a top-level recursion; or else
really belong to some outer recursion universe, if this is an inner
recursion.  Next we have \texttt{fold}, which will allow us to fold
another algebra over data of type \verb|R|.  We will use \verb|fold|
to initiate subsidiary recursions.  Then there is \verb|eval|, which
is used to make recursive calls on data of type \verb|R|.

As noted already, for subsidiary recursion, algebras have a carrier
\verb|X| which depends (functorially) on a type.  This is so that (i)
inside an inner recursion we may compute a result of some type that
may mention \verb|R|, but (ii) outside that recursion, the result will
mention the anchor type \verb|C|.  The \verb|eval| function returns
something of type \verb|X R|, and so does the algebra itself; this
demonstrates (i).  For (ii): if we look at the definition of
\verb|FoldT| in the figure, we see that folding an algebra of type
\verb|alg C X| over a value of type \verb|C| produces a result of type
\verb|X C|.  Having a functor for the carrier of the algebra gives us
the flexibility to type results inside a recursion with the abstract
type \verb|R|, but view those results as having the anchor type
\verb|C| outside the recursion.

@@

@@

As we have seen, our definitions require a form of positive-recursive
types, to allow algebras to accept fold functions that themselves
require algebras, and also for the definition of \verb|Subrec|.  But
as recalled already, full positive-recursive types are incompatible
with Coq's type theory~\cite{coquand88}.  It is worth noting that one
can impose some restrictions on large eliminations which then allow
positive-recursive types~\cite{blanqui05}.  This approach would
require changing the underlying theory.  To avoid this, we here take a
different approach, exploiting Coq's impredicative polymorphism.

This is done in a file \verb|Mu.v|, whose central definitions are in
Figure~\ref{fig:mu}.  The development is parametrized by
\verb|F : Set -> Set| which is assumed to have an \verb|fmap| function
(morphism part of the functor) of type
\begin{verbatim}
forall A B : Set, (A -> B) -> F A -> F B
\end{verbatim}
\noindent which satisfies the identity-preservation law for functors:
\begin{verbatim}
fmapId : forall (A : Set)(d : F A), fmap (fun x => x) d = d
\end{verbatim}

\begin{figure}
\begin{verbatim}
  Inductive Mu : Set := 
    mu : forall (R : Set), (R -> Mu) -> F R -> Mu.

  Definition inMu(d : F Mu) : Mu :=
    mu Mu (fun x => x) d.

  Definition outMu(m : Mu) : F Mu :=
    match m with
    | mu A r d => fmap r d
    end.

  Lemma outIn(d : F Mu) : outMu (inMu d) = d.
\end{verbatim}
\caption{Derivation of retractive-positive recursive types}
\label{fig:mu}
\end{figure}

Let us consider the code in Figure~\ref{fig:mu}.  The critical
idea is embodied in the definition of \verb|Mu|.  We would like
to have a definition like
\begin{verbatim}
  Inductive Mu' : Set := mu' : F Mu' -> Mu'.
\end{verbatim}
\noindent This is exactly what is used in approaches to modular
datatypes in functional programming, like
Swierstra's~\cite{swierstra08}.  But this definition is (rightly)
rejected by Coq, as it would unsoundly enable instantiations of
\verb|F| that are not strictly positive.

Instead, the definition of \verb|Mu| in Figure~\ref{fig:mu} weakens
this ideal definition to a strictly positive approximation:
\begin{verbatim}
  Inductive Mu : Set := 
    mu : forall (R : Set), (R -> Mu) -> F R -> Mu.
\end{verbatim}
\noindent Instead of taking in \verb|F Mu|, constructor \verb|mu|
accepts an input of type \verb|F R|, for some type \verb|R| for which
we have a function of type \verb|R -> Mu|.  The impredicative
quantification of \verb|R| is essential here: we instantiate it with
\verb|Mu| itself in the definition of \verb|inMu|
(Figure~\ref{fig:mu}).  So this approach would not work in a
predicative theory like Agda's.  The quantification of \verb|R| can be
seen as applying a technique due to Mendler, of introducing
universally quantified variables for problematic type occurrences, to
a datatype constructor.  We will review this in
Section~\ref{sec:related}.

Returning to Figure~\ref{fig:mu}, we have functions \verb|inMu| and
\verb|outMu|, which make \verb|F Mu| a retraction (\verb|outIn|) of
\verb|Mu|: the composition of \verb|outMu| and \verb|inMu| is
(extensionally) the identity on \verb|F Mu|.  But the reverse
composition cannot be proved to be the identity, because of the basic
problem of \textbf{noncanonicity} that arises with this definition.

For a simple example of noncanonicity, suppose we instantiate \verb|F|
with \verb|ListF| (of Figure~\ref{fig:listf}).  Please note that as
\verb|Mu| is used in our derivation of subsidiary recursion, we will
not instantiate this \verb|F| with the signature functor of a datatype
directly; but this will show the issue in a simple form.  Let us
temporarily define \verb|List A| as \verb|Mu (ListF A)| (again, for
subsidiary recursion we use a different functor than just \verb|ListF|
directly).  The canonical way to define the empty list would be, 
implicitly instantiating \verb|F| to \verb|ListF A|,
\begin{verbatim}
Definition mkNil := mu (List A) (fun x => x) (NilF A)
\end{verbatim}
\noindent But given this, there are infinitely many other equivalent
definitions.  For any \verb|Q : Set|, we could take
\begin{verbatim}
Definition mkNil' := mu Q (fun x => mkNil) (NilF A)
\end{verbatim}
\noindent Since \verb|fmap f (NilF A)| equals just \verb|NilF B| for
\verb|f : A -> B|, if we apply \verb|outMu| (of Figure~\ref{fig:mu})
to \verb|mkNil'| or \verb|mkNil|, we will get \verb|NilF (List A)|.
But critically, \verb|mkNil| and \verb|mkNil'| are not equal, neither
definitionally nor provably.  One can define a function that puts
\verb|Mu| values in normal form by folding \verb|inMu| over them.
Then \verb|mkNil| and \verb|mkNil'| will have the same normal form,
and be equivalent in that sense.  But the fact that they are not
provably equal is what we term noncanonicity.

Noncanonicity leads to some issues, as we turn next to the problem of
inductive reasoning about subsidiary recursions.  With some care,
however, we can avoid pitfalls, leaving us with a form of
positive-recursive type that enables our definitions to go through.

\section{Examples of subsidiary induction}
\label{sec:examplesi}

@@

\subsection{Termination}
In some tools, like Coq, Agda, and Lean, termination is checked
statically, based on structural decrease at recursive calls.  Others,
like Isabelle/HOL, allow one to write recursions first, and prove
(possibly with automated help) their termination
afterwards~\cite{krauss}.

@@

Subsidiary recursion can be seen as a generalization of \emph{nested
recursion}, which allows recursive calls of the form \verb|f (f x)|.
In subsidiary recursion, these are generalized to the form
\verb|f (g x)|, where \verb|g| could be \verb|f| or another
recursively defined function.

\subsection{Mendler encoding}

Mendler introduced the basic idea of using universal abstraction to
support compositional termination checking; an accessible source
is~\cite{mendler91}.  This recursor has type
\[
\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to X}
\]
\noindent We have adopted this idea to the constructor of the type
\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
categorical perspective on Mendler-style recursion~\cite{uustalu99}.
Others have explored the possibility of using it with negative type
schemes~\cite{ahn11}.
